---
title: "Getting Started With rExpertQuery"
format: html
editor: visual
author: "ATTAINS Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 8
    fig_width: 8
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting Started With rExpertQuery}
  %\VignetteEngine{knitr::rmarkdown}
description: An overview of rExpertQuery functions for querying and downloading ATTAINS data via Expert Query web services.
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
library(knitr)
library(dplyr)
library(DT)
```

```{css, echo = F, eval = FALSE}
pre {
  max-height: auto;
  overflow-y: auto;
}

pre[class] {
  max-height: auto;
}
```

## About Expert Query and rExpertQuery

Expery Query (https://owapps.epa.gov/expertquery/attains/tmdl) empowers users to access surface 
water quality data from the Assessment and TMDL Tracking and Implementation System (ATTAINS), 
encompassing Assessment decisions (under Clean Water Act Sections 303(d), 305(b), and 106) and 
Action data like Total Maximum Daily Loads (TMDLs), Advance Restoration Plans (ARPs), and 
Protection Approaches.

The tool supports querying and downloading data within and across organizations (states or tribes), 
such as querying all the data within an EPA region or all nutrient TMDLs nationally. It also allows 
users to download national ATTAINS data extracts.

rExpertQuery takes this a step further as its functions facilitate querying the Expert Query web
services from R functions which allow the user to import the resulting data sets directly in to
R for more detailed review and analysis.

It is crucial to be aware of certain complexities inherent in the data and querying process. Users 
are cautioned against direct comparisons between states due to differences in water quality standards 
and assessment methodologies. 

While Expert Query facilitates data querying, it lacks built-in summary capabilities. Users are 
encouraged to download and summarize data, but use caution when relying on row counts, as duplicate 
entries may occur. This complexity arises from multiple tables that have many-to-many relationships.
Future vignettes and functions may provide examples of how to summarize Expert Query data.


## Installation

You must first have R and R Studio installed to use the rExpert Query (see instructions below if 
needed). rExpert Query is in active development, therefore we highly recommend that you update it
and all of its dependency libraries each time you use the package. 

You can install and/or update the [rExpertQuery Package](https://github.com/USEPA/rExpertQuery) and 
all dependencies by running:

```{r install, eval = TRUE, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE}
if (!"remotes" %in% installed.packages()) {
  install.packages("remotes")
}

remotes::install_github("USEPA/rExpertQuery", ref = "develop", dependencies = TRUE, force = TRUE)

library(dplyr)
library(data.table)
```

```{r testkey, include=FALSE}
testkey <- "53BVce47MQ3KXKibjx35g4ojaDQGh8qWfbdO8cE0"
```


## EQ_Actions

*EQ_Actions()* queries ATTAINS Actions data by a variety of user supplied parameters. For example,
you might like to run a relatively simple query  return data for all Region 4 Actions that should be
included in Measures.

```{r eq.actions.r4, echo = TRUE, results = FALSE, message = FALSE, warning = FALSE}
# query Actions from Region 4 that are included in Measures
R4_actions_in_meas <- rExpertQuery::EQ_Actions(api_key = testkey, region = 4, in_meas = "Yes")
```

This query returns `r formatC(nrow(R4_actions_in_meas), big.mark = ",")` results, so we'll review only a small random subset of them in a data 
table just to get an idea of what those results look like:

```{r r4.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
R4_actions_in_meas$planSummaryLink <- paste0('<a href="', R4_actions_in_meas$planSummaryLink, '" target="_blank">', R4_actions_in_meas$planSummaryLink, "</a>")
```

```{r r4.datatable}
# create random subset of R4 query results
R4_subset <- R4_actions_in_meas %>%
  dplyr::slice_sample(n = 20)

# create data tab;e
DT::datatable(R4_subset, options = list(pageLength = 2, scrollX = TRUE))
```

You might also want to create a more specific query. For example, you might want to find all 
Actions from Missouri that were issued by EPA with completion dates between 2000-01-01 and 
2020-12-31. As this more detailed query returns a smaller data set, the data table shows all
results of the query

```{r eq.actions.ex2, results = TRUE, message = FALSE, warning = FALSE}
# query Actions for Missouri Actions with action agency "EPA" and parameter group "PATHOGENS"
MO_epa <- rExpertQuery::EQ_Actions(
  api_key = testkey, statecode = "MO",
  act_agency = "EPA",
  param_group = "PATHOGENS",
  comp_date_start = "2000-01-01",
  comp_date_end = "2020-12-31"
)
```

```{r mo.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
MO_epa$planSummaryLink <- paste0('<a href="', MO_epa$planSummaryLink, '" target="_blank">', MO_epa$planSummaryLink, "</a>")
```

```{r mo.data.table}
# view MO results
DT::datatable(MO_epa, options = list(pageLength = 10, scrollX = TRUE))
```
## EQ_ActionsDocuments

*EQ_Actions()* queries ATTAINS Actions Documents data with user supplied parameters. One
of its unique features is that you can search the documents themselves by keyword or phrase. For
example, you might want to find all Actions Documents which contained the keyword "nutrients".

```{r nutrients, results = TRUE, message = FALSE, warning = FALSE}
# Actions Documents containing "nutrient"
Nutrient_docs <- rExpertQuery::EQ_ActionsDocuments(
  doc_query = "nutrient",
  api_key = testkey
)
```

This query yields `r formatC(nrow(Nutrient_docs), big.mark = ",")` results. As well as providing the Action and Document Name for all results, *EQ_Actions()* also returns the column "actionDocumentUrl"
containg the URL to link to the document. Becaue the query yields so many results, we'll take a look at a smaller subset of them in a data table to understand the structure of the results.

```{r nutrients.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
Nutrient_docs$actionDocumentUrl <- paste0('<a href="', Nutrient_docs$actionDocumentUrl, '" target="_blank">', Nutrient_docs$actionDocumentUrl, "</a>")
```

```{r nutrients.data.table}
Nutrient_docs_subset <- Nutrient_docs %>%
  dplyr::slice_sample(n = 20)

DT::datatable(Nutrient_docs_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

You can also query by more typical parameters like organization name (org_name), EPA region (region)
, or action type (act_type). The example below demonstrates using the function to search for 4B Restoration Approach documents from Pennsylvania.

```{r pa.4b.ex, results = TRUE, message = FALSE, warning = FALSE}
PA_4B <- rExpertQuery::EQ_ActionsDocuments(
  statecode = "PA", act_type = "4B Restoration Approach",
  api = testkey
)
```

```{r pa.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
PA_4B$actionDocumentUrl <- paste0('<a href="', PA_4B$actionDocumentUrl, '" target="_blank">', PA_4B$actionDocumentUrl, "</a>")
```
 
```{r pa.4b.table}
PA_4B_subset <- PA_4B %>%
  dplyr::slice_sample(n = 20)

DT::datatable(PA_4B_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```
 
 
 As there are only `r formatC(nrow(PA_4B), big.mark = ",")` 4B Restoration Approach documents from Pennsylvania, we can review them all in the table below. As in the other Actions Document Example,
 the actionDocumentUrl column contains a link to the document.

## EQ_Assessments
*EQ_Assessments()* queries ATTAINS Assessments data. For many use cases, querying Assessments for
a single organization or state is desirable. The next example shows how to query by a single state and use class for the latest cycle.
```{r ky.assess, results = TRUE, message = FALSE, warning = FALSE}
KY_assessments <- rExpertQuery::EQ_Assessments(
  statecode = "KY",
  api_key = testkey
)
```

There are `r formatC(nrow(KY_assessments), big.mark = ",")` results for the basic latest cycle Kentucky query. It is also possible to design more specific assessment queries. 

For example, you might want to narrow down the results to show causes for a use class in one 
region. The code below demonstrates querying for EPA Region 3 Assessments in the ECOLOGICAL_USE 
group with at least one parameter with the status "Cause". The data table below shows a random 
sample of the results from this Region 3 Query.

```{r R3.ecological, results = TRUE, message = FALSE, warning = FALSE}
R3_ecological <- rExpertQuery::EQ_Assessments(
  region = 3,
  use_group = "ECOLOGICAL_USE",
  param_status = "Cause",
  api_key = testkey
)
```

```{r r3.ecological.table}
R3_ecological_subset <- R3_ecological %>%
  dplyr::slice_sample(n = 20)

DT::datatable(R3_ecological_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```
## EQ_AssessmentUnits

*EQ_AssessmentUnits()* queries ATTAINS Assessment Units data. The default is to only return "Active"
assessment units, but input params can be adjusted to query for "Retired" or "Historical"
assessment units in addition to or instead of "Active". The example below shows a simple example for
querying for Active assessment units for a single state.

```{r ri.aus}

RI_aus <- rExpertQuery::EQ_AssessmentUnits(statecode = "RI",
                                           api_key = testkey)

```

```{r ar.retired}

AR_retired <- rExpertQuery::EQ_AssessmentUnits(statecode = "AR",
                                           au_status = "Retired",
                                           api_key = testkey)

```

## EQ_AUsMLs

Content in development
```{r}
AK_ausmls <- rExpertQuery::EQ_AssessmentUnits(statecode = "AK",
                                              api_key = testkey)
```

```{r}
WA_ausmls <- rExpertQuery::EQ_AssessmentUnits(statecode = "WA",
                                              api_key = testkey)
```

## EQ_CatchCorr

Content in development
```{r dc.catchcorr}

DC_catch <- rExpertQuery::EQ_CatchCorr(statecode = "DC")

```

```{r il.au.catch}

IL_dj12_catch <- rExpertQuery::EQ_CatchCorr(statecode = "IL",
                          auid = "IL_DJ-12",
                          api_key = testkey)

```

## EQ_NationalExtract

```{r nat.tmdls}

Nat_tmdls <- rExpertQuery::EQ_NationalExtract("tmdls")

```

```{r nat.tmdls}

Nat_actions <- rExpertQuery::EQ_NationalExtract("actions")

```
## EQ_Sources

Content in development
```{r wi.source.hab}
WI_habalt_sources <- rExpertQuery::EQ_Sources(statecode = "WI",
                                 param_group = "HABITAT ALTERATIONS",
                                 api_key = testkey)
```

```{r source.legacy}
legacy_sources <- EQ_Sources(source = "LEGACY/HISTORICAL POLLUTANTS",
                             api_key = testkey)
```

## EQ_TMDLs

Content in development
```{r hi.tmdl.both}

HI_both_tmdls <- rExpertQuery::EQ_TMDLs(statecode = "HI",
                               source_type = "Both",
                               api_key = testkey)
```

```{r pathogens.r10}

R10_pathogens <- rExpertQuery::EQ_TMDLs(region = 10,
                             ad_param_group = "PATHOGENS",
                             api_key = testkey)
               
```


## EQ_DomainValues

Content in development
```{r see.dom.vals}

Dom_vals <- rExpertQuery::


```



