---
title: "Getting Started With rExpertQuery"
format: html
editor: visual
author: "ATTAINS Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 8
    fig_width: 8
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting Started With rExpertQuery}
  %\VignetteEngine{knitr::rmarkdown}
description: An overview of using rExpertQuery functions to retrieve ATTAINS data via Expert Query
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---
---

```{r setup, include = F}
library(knitr)

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

```{css, include = F}
pre {
  max-height: 300px;
  overflow-y: auto;
}

pre[class] {
  max-height: 300px;
}
```

## rExpert Query

rExpert Query is an R package designed to help users easily access Assessment and TMDL Tracking and Implementation System (ATTAINS) data via Expert Query web services. Expert Query empowers users to access surface water quality data from ATTAINS encompassing Assessment decisions (under Clean Water 
Act Sections 303(d), 305(b), and 106) and Action data like Total Maximum Daily Loads (TMDLs), 
Advance Restoration Plans (ARPs), and Protection Approaches.

rExpertQuery supports querying and downloading data within and across organizations (states or 
tribes), such as querying all the data within an EPA region or all nutrient TMDLs nationally. 
However, it's crucial to be aware of certain complexities inherent in the data and the querying process. Users are cautioned against direct comparisons between states due to differences in water quality standards and assessment methodologies. 

While Expert Query facilitates data querying, it lacks built-in summary capabilities. Users are encouraged to download and summarize data, but use caution when relying on row counts, as duplicate entries may occur. The complexity arises from multiple tables that have many-to-many relationships. 

## Expert Query API Keys

rExpertQuery functions require the use of an API key. The exception is the EQ_NationalExtract
function which downloads a copy of the national extract for the selected profile. If you only want
to download national extracts, you do not need an API key. If you wish to use any other 
rExpertQuery function you must visit: https://owapps.epa.gov/expertquery/api-key-signup to
sign up for your own free API key.

The examples in this vignette are running using a test key. Do not use the test key in your own workflows.

## EQ_Actions

*EQ_Actions* queries ATTAINS Actions data via Expert Query web services and returns the results as 
a data frame. Commonly used query paramters include: param_name (Parameter Name), 
org_id (Oranization Identifier), and water_type (Waterbody Type). A full list of the query 
parameters available can be found with "?EQ_Actions".

```{r EQ_Actions doc}
# see EQ_Actions documentation
?EQ_Actions
```








