---
title: "Getting Started With rExpertQuery"
format: html
editor: visual
author: "ATTAINS Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_caption: yes
    fig_height: 8
    fig_width: 8
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Getting Started With rExpertQuery}
  %\VignetteEngine{knitr::rmarkdown}
description: An overview of rExpertQuery functions for querying and downloading ATTAINS data via Expert Query web services.
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
library(knitr)
library(dplyr)
library(DT)
```

```{css, echo = F, eval = FALSE}
pre {
  max-height: auto;
  overflow-y: auto;
}

pre[class] {
  max-height: auto;
}
```

## About Expert Query and rExpertQuery

[Expert Query](https://owapps.epa.gov/expertquery/attains/tmdl) empowers
users to access surface water quality data from the [Assessment and TMDL
Tracking and Implementation System
(ATTAINS)](https://www.epa.gov/waterdata/upload-data-resources-registered-attains-usershttps://www.epa.gov/waterdata/upload-data-resources-registered-attains-users)
encompassing Assessment decisions (under Clean Water Act Sections
303(d), 305(b), and 106) and Action data like Total Maximum Daily Loads
(TMDLs), Advance Restoration Plans (ARPs), and Protection Approaches.

The tool supports querying and downloading data within and across
organizations (states or tribes), such as querying all the data within
an EPA region or all nutrient TMDLs nationally. It also allows users to
download national ATTAINS data extracts.

rExpertQuery takes this a step further as its functions facilitate
querying the Expert Query web services from R functions which allow the
user to import the resulting data sets directly in to R for more
detailed review and analysis.

It is crucial to be aware of certain complexities inherent in the data
and querying process. Users are cautioned against direct comparisons
between states due to differences in water quality standards and
assessment methodologies.

While Expert Query facilitates data querying, it lacks built-in summary
capabilities. Users are encouraged to download and summarize data, but
use caution when relying on row counts, as duplicate entries may occur.
This complexity arises from multiple tables that have many-to-many
relationships. Future vignettes and functions may provide examples of
how to summarize Expert Query data.

## Installation

You must first have R and R Studio installed to use the rExpert Query
(see instructions below if needed). rExpert Query is in active
development, therefore we highly recommend that you update it and all of
its dependency libraries each time you use the package.

You can install and/or update the [rExpertQuery
Package](https://github.com/USEPA/rExpertQuery) and all dependencies by
running:

```{r install, eval = TRUE, echo = TRUE, results = 'hide', message = FALSE, warning = FALSE}
if (!"remotes" %in% installed.packages()) {
  install.packages("remotes")
}

remotes::install_github("USEPA/rExpertQuery", ref = "develop", dependencies = TRUE, force = TRUE)

library(dplyr)
library(data.table)
```

```{r testkey, include=FALSE}
testkey <- "53BVce47MQ3KXKibjx35g4ojaDQGh8qWfbdO8cE0"
```

## EQ_Actions

[**EQ_Actions()**](https://usepa.github.io/rExpertQuery/reference/EQ_Actions.htmlhttps://usepa.github.io/rExpertQuery/reference/EQ_Actions.html)queries
ATTAINS Actions data by a variety of user supplied parameters. For
example, you might like to run a relatively simple query return data for
all Region 4 Actions that should not be included in Measures.

```{r eq.actions.r4, echo = TRUE, results = FALSE, message = FALSE, warning = FALSE}
# query Actions from Region 4 that are not included in Measures
R4_actions_not_meas <- rExpertQuery::EQ_Actions(api_key = testkey, region = 4, in_meas = "No")
```

This query returns
`r formatC(nrow(R4_actions_not_meas), big.mark = ",")` results, so we'll
review only a small random subset of them in a data table just to get an
idea of what those results look like:

```{r r4.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
R4_actions_not_meas$planSummaryLink <- paste0('<a href="', R4_actions_not_meas$planSummaryLink, '" target="_blank">', R4_actions_not_meas$planSummaryLink, "</a>")
```

```{r r4.datatable}
# create random subset of R4 query results
R4_subset <- R4_actions_not_meas %>%
  dplyr::slice_sample(n = 20)

# create data tab;e
DT::datatable(R4_subset, options = list(pageLength = 2, scrollX = TRUE))
```

You might also want to create a more specific query. For example, you
might want to find all Actions from Missouri that were issued by EPA
with completion dates between 2000-01-01 and 2020-12-31. As this more
detailed query returns a smaller data set, the data table shows all
results of the query

```{r eq.actions.ex2, results = TRUE, message = FALSE, warning = FALSE}
# query Actions for Missouri Actions with action agency "EPA" and parameter group "PATHOGENS"
MO_epa <- rExpertQuery::EQ_Actions(
  api_key = testkey, statecode = "MO",
  act_agency = "EPA",
  param_group = "PATHOGENS",
  comp_date_start = "2000-01-01",
  comp_date_end = "2020-12-31"
)
```

```{r mo.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
MO_epa$planSummaryLink <- paste0('<a href="', MO_epa$planSummaryLink, '" target="_blank">', MO_epa$planSummaryLink, "</a>")
```

```{r mo.data.table}
# view MO results
DT::datatable(MO_epa, options = list(pageLength = 10, scrollX = TRUE))
```

## EQ_ActionsDocuments

[**EQ_ActionsDocuments()**](https://usepa.github.io/rExpertQuery/reference/EQ_ActionsDocuments.html)queries
ATTAINS Actions Documents data with user supplied parameters. One of its
unique features is that you can search the documents themselves by
keyword or phrase. For example, you might want to find all Actions
Documents which contained the keyword "nutrients".

```{r nutrients, results = TRUE, message = FALSE, warning = FALSE}
# Actions Documents containing "nutrient"
Nutrient_docs <- rExpertQuery::EQ_ActionsDocuments(
  doc_query = "nutrient",
  api_key = testkey
)
```

This query yields `r formatC(nrow(Nutrient_docs), big.mark = ",")`
results. As well as providing the Action and Document Name for all
results, *EQ_Actions()* also returns the column "actionDocumentUrl"
containg the URL to link to the document. Becaue the query yields so
many results, we'll take a look at a smaller subset of them in a data
table to understand the structure of the results.

```{r nutrients.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
Nutrient_docs$actionDocumentUrl <- paste0('<a href="', Nutrient_docs$actionDocumentUrl, '" target="_blank">', Nutrient_docs$actionDocumentUrl, "</a>")
```

```{r nutrients.data.table}
Nutrient_docs_subset <- Nutrient_docs %>%
  dplyr::slice_sample(n = 20)

DT::datatable(Nutrient_docs_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

You can also query by more typical parameters like organization name
(org_name), EPA region (region) , or action type (act_type). The example
below demonstrates using the function to search for 4B Restoration
Approach documents from Pennsylvania.

```{r pa.4b.ex, results = TRUE, message = FALSE, warning = FALSE}
PA_4B <- rExpertQuery::EQ_ActionsDocuments(
  statecode = "PA", act_type = "4B Restoration Approach",
  api = testkey
)
```

```{r pa.url.prep, include = FALSE}
# Actions Documents containing "nutrient"
PA_4B$actionDocumentUrl <- paste0('<a href="', PA_4B$actionDocumentUrl, '" target="_blank">', PA_4B$actionDocumentUrl, "</a>")
```

```{r pa.4b.table}
PA_4B_subset <- PA_4B %>%
  dplyr::slice_sample(n = 20)

DT::datatable(PA_4B_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

As there are only `r formatC(nrow(PA_4B), big.mark = ",")` 4B
Restoration Approach documents from Pennsylvania, we can review them all
in the table below. As in the other Actions Document Example, the
actionDocumentUrl column contains a link to the document.

## EQ_Assessments

[**EQ_Assessments()**](https://usepa.github.io/rExpertQuery/reference/EQ_Assessments.html)queries
ATTAINS Assessments data. For many use cases, querying Assessments for a
single organization or state is desirable. The next example shows how to
query by a single state and use class for the latest cycle.

```{r ky.assess, results = TRUE, message = FALSE, warning = FALSE}
KY_assessments <- rExpertQuery::EQ_Assessments(
  statecode = "KY",
  api_key = testkey
)
```

There are `r formatC(nrow(KY_assessments), big.mark = ",")` results for
the basic latest cycle Kentucky query.

```{r ky.table}
KY_subset <- KY_assessments %>%
  dplyr::slice_sample(n = 20)

DT::datatable(KY_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)

```

It is also possible to design more specific assessment queries. For
example, you might want to narrow down the results to show causes for a
use class in one region. The code below demonstrates querying for EPA
Region 3 Assessments in the ECOLOGICAL_USE group with at least one
parameter with the status "Cause". The data table below shows a random
sample of the results from this Region 3 Query.

```{r R3.ecological, results = TRUE, message = FALSE, warning = FALSE}
R3_ecological <- rExpertQuery::EQ_Assessments(
  region = 3,
  use_group = "ECOLOGICAL_USE",
  param_status = "Cause",
  api_key = testkey
)
```

```{r r3.ecological.table}
R3_ecological_subset <- R3_ecological %>%
  dplyr::slice_sample(n = 20)

DT::datatable(R3_ecological_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

## EQ_AssessmentUnits

[**EQ_AssessmentUnits()**](https://usepa.github.io/rExpertQuery/reference/EQ_AssessmentUnits.html)queries
ATTAINS Assessment Units data. The default is to only return "Active"
assessment units, but input params can be adjusted to query for
"Retired" or "Historical" assessment units in addition to or instead of
"Active". The example below shows a simple example for querying for
Active assessment units for a single state.

```{r ri.aus, results = TRUE, message = FALSE, warning = FALSE}

RI_aus <- rExpertQuery::EQ_AssessmentUnits(statecode = "RI",
                                           api_key = testkey)

```

We can take a look at a random selection of 20 of the active assessment
units from Rhode Island.

```{r ri.table}

RI_aus_subset <- RI_aus %>%
  dplyr::slice_sample(n = 20)

DT::datatable(RI_aus_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

It is is also possible to search fo retired assessment units, as show in
the example from Florida below.

```{r fl.retired, results = TRUE, message = FALSE, warning = FALSE}

FL_retired <- rExpertQuery::EQ_AssessmentUnits(statecode = "FL",
                                           au_status = "Retired",
                                           api_key = testkey)

```

```{r ri.table}

FL_ret_subset <- FL_retired %>%
  dplyr::slice_sample(n = 20)

DT::datatable(FL_ret_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

## EQ_AUsMLs

[**EQ_AUsMLs()**](https://usepa.github.io/rExpertQuery/reference/EQ_AUsMLs.html)returns
information on the monitoring locations used to make assessment
determinations at specific assessment units. Not all organizations
provide monitoring location data in ATTAINS. The first example uses a
state that does supply some monitoring location data.

```{r akausmls, results = TRUE, message = FALSE, warning = FALSE}
AK_ausmls <- rExpertQuery::EQ_AUsMLs(statecode = "AK",
                                              api_key = testkey)

AK_filt <- AK_ausmls %>% dplyr::filter(!is.na(monitoringLocationId))
```

While the resulting data set from the query has
`r formatC(nrow(AK_ausmls), big.mark = ",")` results, filtering the data
set to retain only records where a monitoring location id has been
recorded leads to a data set with
`r formatC(nrow(AK_filt), big.mark = ",")` results.

## EQ_CatchCorr

[**EQ_CatchCorr()**](https://usepa.github.io/rExpertQuery/reference/EQ_CatchCorr.html)queries
ATTAINS Catchment Correspondance data. These queries return large files,
particularly if a large geospatial area is defined, so best practice is
to make your query as specific as possible. Many machines may not have
enough memory to load an entire state's worth of catchment
correspondence data in an R session.

```{r dc.catchcorr, results = TRUE, message = TRUE, warning = FALSE}

DC_catch <- rExpertQuery::EQ_CatchCorr(statecode = "DC",
                                       api_key = testkey)

```

The table below shows 20 random results from the DC catchment
correspondence query.

```{r dc.table}

DC_subset <- DC_catch %>%
  dplyr::slice_sample(n = 20)

DT::datatable(DC_subset, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)
```

It is also possible to focus on a specific assessment unit in catchment
correspondence queries. The next example queries for a single Illinois
assessment unit by its id.

```{r il.au.catch, results = TRUE, message = FALSE, warning = FALSE}

IL_N99_catch <- rExpertQuery::EQ_CatchCorr(statecode = "IL",
                          auid = "IL_N-99",
                          api_key = testkey)

```

The resulting table shows all results from this much smaller query.

```{r il.table}

IL_N99_subset <- IL_N99_catch %>%
  dplyr::slice_sample(n = 20)

DT::datatable(IL_N99_catch, options = list(pageLength = 5, scrollX = TRUE), escape = FALSE)

```

## EQ_NationalExtract

[**EQ_NationalExtract()**](https://usepa.github.io/rExpertQuery/reference/EQ_NationalExtract.html)provides
an efficient method for importing the [Expert Query National
Extracts](https://owapps.epa.gov/expertquery/national-downloads). This
function requires one argument, "extract" to specify which of the
National Extracts to download. The following examples show how to
download the TMDL and Actions extracts. The extracts are all large files
and some may not open in R if there is not enough memory available.

```{r nat.tmdls, results = TRUE, message = FALSE, warning = FALSE}

Nat_tmdls <- rExpertQuery::EQ_NationalExtract("tmdl")

```

```{r nat.actions, results = TRUE, message = FALSE, warning = FALSE}

Nat_actions <- rExpertQuery::EQ_NationalExtract("actions")

```

## EQ_Sources

[**EQ_Sources()**](https://usepa.github.io/rExpertQuery/reference/EQ_Sources.html)

Additional content in development

```{r wi.source.hab, results = TRUE, message = FALSE, warning = FALSE}
WI_habalt_sources <- rExpertQuery::EQ_Sources(statecode = "WI",
                                 param_group = "HABITAT ALTERATIONS",
                                 api_key = testkey)
```

```{r source.legacy, results = TRUE, message = FALSE, warning = FALSE}
legacy_sources <- rExpertQuery::EQ_Sources(source = "LEGACY/HISTORICAL POLLUTANTS",
                             api_key = testkey)
```

## EQ_TMDLs

[**EQ_TMDLs()**](https://usepa.github.io/rExpertQuery/reference/EQ_TMDLs.html)

Additional content in development

```{r hi.tmdl.both, results = TRUE, message = FALSE, warning = FALSE}

HI_both_tmdls <- rExpertQuery::EQ_TMDLs(statecode = "HI",
                               source_type = "Both",
                               api_key = testkey)
```

```{r unknown.r10, results = TRUE, message = FALSE, warning = FALSE}

# R10_unknown <- rExpertQuery::EQ_TMDLs(region = 10,
#                              ad_param_group = "CAUSE UNKNOWN",
#                              api_key = testkey)
#                
```

## EQ_DomainValues

[**EQ_DomainValues()**](https://usepa.github.io/rExpertQuery/reference/EQ_DomainValues.html)

Additional content in development

```{r see.dom.vals, results = TRUE, message = FALSE, warning = FALSE}

# Dom_vals <- rExpertQuery::EQ_DomainValues()


```

```{r dom.vals.org, results = TRUE, message = FALSE, warning = FALSE}

# Org_vals <- rExpertQuery::EQ_DomainValues(domain = "org_id")

```
